### 2023-02-28 intro.tex
Add new constraint rules using marked expression syntax and narrative surrounding them. Removes old rules (some errors due to stray references to them persist)

Contemporary programming environments provide 
a variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g. semantic navigation, type hints, semantic code completion, and automated refactoring. However, during the editing process,
it is common for programs to be ill-typed.

Language designers typically specify static type systems as a collection of rules 
stating conditions under which an expression can be judged to be well-typed,
i.e. completely free of type errors. 
 is well-supported 
by these definitions.

When there is a type error, however, these definitions are inadequate. It is therefore left as an implementation decision (1) type error localization: how to localize and explain the type error, and (2) type error recovery: how to recover from the error to find other simultaneous errors and to continue to provision various semantic services, which would often still useful even in the presence of errors.

Consider, for example,

let x = 
  if f(y) then 1 else true
in 
  x + 1

(compare Elm/OCaml/Haskell and Hazel directly?)

Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.

For these reasons, accurate type error localization and robust type error recovery have been a topic of considerable practical interest. 
Indeed, essentially all widely-used programming environments (often via an underlying language server) have some support for type error 
localization and many also attempt recovery of certain services in certain situations. However, the design of these systems has evolved in some part as folklore, as type system definitions have remained silent.

One reason for this is 
that type system definitions typically only define the conditions under which an expression is well-typed, e.g. in the style of the simply typed lambda calculus and its derivatives. When there is a type error anywhere in the program, it is everywhere meaningless according to definitions in this style, i.e. there is no direct specification for how to reason within an erroneous program.

The primary contribution of this paper is to provide these missing theoretical foundations for type error localization and recovery. 
Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 
(Mention gradual typing here?)

>>>>

Modern programming environments provide developers with 
various semantic services, such as type hints, semantic navigation, semantic code completion, and automated refactoring, that require statically reasoning about the type and binding structure of the program as it is being edited. 
% These semantic services are often packaged into language servers\todo{cites}.
The problem is that the program being edited can be ill-typed, sometimes for long periods of time. 
For example, when the developer changes an important type definition, errors might arise at dozens of usage sites in a large program, which could take hours or days to address. Semantic services often exhibit gaps in service during these periods \cite{HazelnutSNAPL}, because in the conventional formal account of type systems, only well-typed programs have a well-defined type and binding structure. As soon as a type error occurs \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

Developers benefit when these gaps in service are minimized or eliminated, so there has long been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for directing the developer to 
the location(s) in a program that may be erroneous, and (2) \textbf{type error recovery}: 
mechanisms that allow the language server to optimistically recover from localized errors and continue to provide downstream semantic services, ideally throughout the program.

Essentially all widely-used programming systems have some support for type error 
localization, e.g. via compiler error messages or user interface affordances fed by data from a language server.\todo{cites}
Many also attempt recovery of certain services in certain situations, discussed below. 
However, type error localization and recovery mechanisms has developed idiosyncratically, often as folklore amongst tool designers. Different type checkers or language servers, even for the same language, localize and recover from type errors in different ways, with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

Consider, for example, the following program written in a typed functional language with, for now, only local type inference \cite{pierce}: 
\begin{lstlisting}[numbers=none]
let x = 
  if f(y) then 1 else true
in 
  x + 1
\end{lstlisting}
A type checker with no support for error localization or recovery at all would simply report that this program is ill-typed, because a type cannot be derived for it according to the type system specification, if given in the conventional style (e.g. following the simply typed lambda calculus and its derivatives).

The more common and practical approach is to localize the first error that causes type checking to fail and emit an error message\todo{cites} (which can help, though studies suggest that developers primarily use location information). In this example, such a system might report that the variable \li{f} is unbound and communicate its location, e.g. using editor decorations. 
This is the approach taken in many production type checkers, e.g. XXX.\todo{examples}

A language server with support for type error recovery might continue past this initial error to also report that \li{y} is unbound. To recover further, the system might optimistically ignore the fact that guard of the conditional cannot be confirmed to be a boolean expression and continue into its branches, observing that they have inconsistent types, \li{int} and \li{bool}.

There are several potential approaches to localizing this inconsistency. 
One approach would be to assume, arbitrarily, that the first branch is correct, so the error can be localized to the second branch, or \emph{vice versa}. 
Heuristic approaches have been developed to make choices like these less arbitrary, e.g. by inspecting the recent editing history \todo{cite}, or training a machine learning model \todo{cite}. 
Another approach is to localize the inconsistency between the two branches to the conditional expression itself. This entirely avoids needing to heuristically decide which branch is correct.

This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
If localization assumes that the \li{then} branch is correct, then \li{x : int} and there is no error to report here.
If the \li{else} branch is chosen, then \li{x : bool} and an error can be reported (though this error might mislead the programmer if this earlier localization step was incorrect). 
If instead the inconsistency is localized to the conditional expression as a whole, then it is not obvious which type to assign to \li{x} at all. (Mention gradual typing here?)

This exercise demonstrates that type error localization and recovery mechanisms engage in non-trivial reasoning and make choices that are not fully specified by the type system itself. 
The first contribution of this paper is to address this problem by developing formal foundations for type error localization and recovery. 
Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 
(Mention gradual typing here?)

### 2023-02-28 let.tex
Add new constraint rules using marked expression syntax and narrative surrounding them. Removes old rules (some errors due to stray references to them persist)

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\ctxSynType{\Gamma_{in}}{p}{\tau ; \Gamma_{out}}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}

    \judgment{ }{
      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMPProd}
  \end{mathpar}
  %
  \caption{Matched product type.}
  \label{fig:calculus-matched-product}
\end{figure}

>>>>

%TODO(andrew): justification, examples

We augment our type system with a new variant on the unknown type which can trigger a swtich from analysis to synthesis. The new type $\TUnknownSwitch$ behaves idenitically to $\TUnknown$ with respect to type consistency, the marched arrow and product judgements, and the join metafunction:
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TUnknownSwitch
\end{array}\]

The unmarked language is extended by let expressions with composite patterns:
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots
                               \mid \ELet{\PMV}{\EMV}{\EMV}
\end{array}\]
\[\begin{array}{rrcl}
  \PMName  & \PMV  & \Coloneqq x 
                               \mid \_
                               \mid (p_1,p_2)
                               \mid p : \TMV

\end{array}\]
%
Let expression have the following semantics:
\begin{mathpar}
    \judgment{
     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
     \\
     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
     \\
     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
     \\\\
     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
     \\
     \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
   }{
     \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
   }{USLet}
   \\
   \judgment{
     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
     \\
     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
     \\
     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
     \\\\
     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
     \\
     \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
   }{
     \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
   }{UALet}
\end{mathpar}
Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new SynSwitch type:
\begin{mathpar}
    \judgment{
     \ctxSynType{\ctx}{\EMV}{_}
   }{
     \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
   }{SynSwitch}

\end{mathpar}

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\ctxSynPat{\Gamma}{\PMV}{\TMV}}} Pattern $\PMV$ synthesizes type $\TMV$
  %
  \begin{mathpar}

   \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{USPatWild}

   \judgment{ }{\ctxSynPat{\ctx}{x}{\TUnknownSwitch}}{USPatVar}

   %\judgment{ }{\ctxSynPat{\Gamma}{\hehole{}}{\TUnknownSwitch}}{USPatHole}
   \judgment{
     \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\_}
   }{
        \ctxSynPat{\ctx}{\PMV : \tau}{}{\tau}
    }{USPatAnn}

   \\\\
   \judgment{
     \ctxSynPat{\ctx}{\PMV_1}{\TMV_1} \\
     \ctxSynPat{\ctx}{\PMV_2}{\TMV_2}
   }{
     \ctxSynPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\TPair{\TMV_1}{\TMV_2}}
   }{USPatPair}

   \judgment{
     \ctxSynPat{\ctx}{\PMV{}}{\_}
   }{
        \ctxSynPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\TUnknownSwitch}
    }{USPatErr}

  \end{mathpar}
  %
  \caption{Pattern Synthesis}
  \label{fig:calculus-pattern-synthesis}
\end{figure}

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\ctxAnaPat{\ctx_{in}}{\PMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against type $\TMV$ producing context $\Gamma_{out}$
  %
  \begin{mathpar}
   \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{UAPatWild}

   \judgment{ }{\ctxAnaPat{\Gamma}{x}{\TMV}{\ctx, x:\TMV}}{UAPatVar}

   %\judgment{ }{\ctxAnaPat{\ctx}{\hehole{}}{\tau}}{UAPatHole}{\ctx}
   \judgment{
     \ctxAnaPat{\ctx}{\PMV{}}{\tau'}{\ctx'}
     \\
     \consistent{\tau}{\tau'}
   }{
        \ctxAnaPat{\ctx}{\PMV : \tau'}{\tau}{\ctx'}
    }{UAPatAnn}

   \\\\
   \judgment{
      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
      \ctxAnaPat{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
      \ctxAnaPat{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
    }{
      \ctxAnaPat{\ctx}{\ECPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
    }{UAPatPair}

   \judgment{
     \ctxSynPat{\ctx}{\PMV{}}{\_}
   }{
        \ctxAnaPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\tau}{\ctx}
    }{UAPatErr}
  \end{mathpar}
  %
  \caption{Pattern Analysis}
  \label{fig:calculus-pattern-analysis}
\end{figure}
